Unnecessarily Complicated Research Title

### 2020-10-29 Chapter4/4_1_intro.tex
chapter_4_passed

The remainder of this chapter is structured as follows:  Section \ref{sec:4_5_related} describes related work; Section \ref{sec:4_2_carsharing} describes details of the three car-sharing paradigms; Section \ref{sec:4_3_methodology} discusses the data collection and analysis methodology for all services; Section \ref{sec:4_4_characterization} presents the results of the characterization for each model and the comparison of them, whereas Section \ref{sec:4_6_conclusion} concludes the paper.

>>>>

The remainder of this chapter is structured as follows:  section \ref{sec:4_5_related} describes related work; section \ref{sec:4_2_carsharing} describes details of the three car-sharing paradigms; section \ref{sec:4_3_methodology} discusses the data collection and analysis methodology for all services; section \ref{sec:4_4_characterization} presents the results of the characterization for each model and the comparison of them, whereas section \ref{sec:4_6_conclusion} concludes the chapter.

### 2020-10-29 Chapter4/4_2_background.tex
chapter_4_passed

The car-sharing systems might be implemented according two paradigms: station based and  free-floating. The station-based paradigm can be divided into \textit{one-way services} and \textit{two-way services}. Both, Station-based models require that a user pick up the vehicle she/he will use at a defined parking spot, i.e. charging station for electric vehicles or providers' parking spots. The user, in turn, may leave the vehicle at any of the base stations scattered throughout the service coverage region (i.e., one-way car-sharing service), or she/he may be obliged to return the vehicle to the station of origin (i.e., two-way car-sharing service). On one hand, the two-way model requires simpler logistics and infrastructure compared to other models. Its implementation can be performed faster and at a lower cost. 
On the other hand, the one-way model may be more flexible and cost-efficient to users than a classical rental. For example, in case there is a base station near to the final user destination, she/he may leave the car at the station while performing other tasks. The time the vehicle is parked is not charged, incurring to lower costs to users. However, a parked vehicle may be reserved by another user. 
The free-floating model does not require any fixed station. In other words, users reserve a car, parked into non-reserved spots in the streets. By the end of the use, users may leave vehicles at any location in a predefined area. Notably, free-floating model eliminates the limitations that station-based models hold, making the experience more flexible and closer to private-owned vehicles~\cite{ciari2014modeling}.

Figure~\ref{fig:diagrama} presents an abstract model that describes the possible states of a vehicle in any of the three car-sharing systems.
Intuitively, a car can be in use (i.e., \textit{busy}) or \textit{idle}. A \textit{busy} vehicle is \textit{rented}, meaning that someone is paying for it during this period. On the other hand, \textit{idle} vehicles may be \textit{unavailable} (i.e., during a maintenance process), \textit{available}, which means someone can reserve or it, or \textit{reserved}.

@@

\begin{figure}[tbh]
\centering
\includegraphics[width=1\textwidth]{imagens/diagrama_v4.pdf}
\caption{Possible states of a vehicle in a car-sharing system.}
\label{fig:diagrama}
\end{figure}

>>>>

The car-sharing systems might be implemented according two paradigms: station based and  free-floating. The station-based paradigm can be divided into \textit{one-way services} and \textit{two-way services}. Both, Station-based models require that a user picks up the vehicle she/he will use at a defined parking spot, i.e. charging station for electric vehicles or providers' parking spots. The user, in turn, may leave the vehicle at any of the base stations scattered throughout the service coverage region (i.e., one-way car-sharing service), or she/he may be obliged to return the vehicle to the station of origin (i.e., two-way car-sharing service). On one hand, the two-way model requires simpler logistics and infrastructure compared to other models. Its implementation can be performed faster and at a lower cost. 
On the other hand, the one-way model may be more flexible and cost-efficient to users than a classical rental. For example, in case there is a base station near to the final user destination, she/he may leave the car at the station while performing other tasks. The time the vehicle is parked is not charged, incurring to lower costs to users. However, a parked vehicle may be reserved by another user. 
The free-floating model does not require any fixed station. In other words, users reserve a car, parked into non-reserved spots in the streets. By the end of the use, users may leave vehicles at any location in a predefined area. Notably, free-floating model eliminates the limitations that station-based models hold, making the experience more flexible and closer to private-owned vehicles~\cite{ciari2014modeling}.

Figure~\ref{fig:4_2_diagrama} presents an abstract model that describes the possible states of a vehicle in any of the three car-sharing systems.
Intuitively, a car can be in use (i.e., \textit{busy}) or \textit{idle}. A \textit{busy} vehicle is \textit{rented}, meaning that someone is paying for it during this period. On the other hand, \textit{idle} vehicles may be \textit{unavailable} (i.e., during a maintenance process), \textit{available}, which means someone can reserve or it, or \textit{reserved}.

@@

\begin{figure}[tbh]
\centering
\includegraphics[width=1\textwidth]{imagens/diagrama_v4.pdf}
\caption{Possible states of a vehicle in a car-sharing system.}
\label{fig:4_2_diagrama}
\end{figure}

### 2020-10-29 Chapter4/4_3_data_collection.tex
chapter_4_passed

%MODO data collection
The Modo data collection process was conducted with a crawler that uses its public API. 
First, we request to the Modo API the list of all vehicles of the service. 
Then, minute by minute, we request the status of each of these vehicles. Each request returns the schedule of a vehicle, informing the periods it will be available for the next 24-hours. Moreover, it returns the vehicle location, i.e., the station with its identifier. 
Note that Modo API does not return specific vehicle status, nor any information that could be used to identify users of the system. We uncover if a vehicle is busy or idle based on its reservation period and the current observation time. In other words, we collect several vehicle schedules and compare each other. Figure~\ref{fig:capturas} illustrates the process of collecting data for a given vehicle. Each data sample corresponds to a request to the API in the order they occur. Data sample \#1 is the result of the API request at minute 1 ($t=1$), data sample \#2 is the result of the API request at minute 2 ($t=2$), etc. At each data sample, the blue dot represents the time a vehicle will be available. We highlight three possible situations:

\begin{itemize}
\item First, as shown in Figure~\ref{fig:capturas}(a), at $t=1$ a given vehicle is shown reserved up to $t=5$.  At $t=2$, the new request to the Modo API still show us that the vehicle will be available only at $t=5$. Each of the following requests to the API confirms the booking period. At the time $t=6$, we perform a request to the API and the vehicle is no longer booked. In sum, we are able to infer that someone booked the vehicle before or at $t=1$, and returned it to the station at $t=5$.

\item Second, as shown in Figure~\ref{fig:capturas}(b), at $t=1$  the Modo API returns that a given vehicle is reserved up to $t=6$.  However, in this case, a request at $t=5$ shows the vehicle is no longer reserved. In this case, we can infer that the user returned the vehicle earlier to the station which means she/he used the vehicle only up to $t=5$.

\item Finally, as shown in Figure~\ref{fig:capturas}(c), the user may extend the booking period. More precisely, at $t=1$ the given vehicle is reserved up to $t=5$. At the third request, we note that the vehicle will no longer be available at $t=5$ but $t=6$. The following API requests confirm the use of the car until $t=6$.
\end{itemize}

@@

    \begin{minipage}[b]{0.32\columnwidth}
    \centering
     \includegraphics[width=\linewidth]{modo_methodology/Normal_Situation.pdf}
     {\\(a) Normal \\situation}
    \end{minipage}
    % \hspace{5mm}
    \begin{minipage}[b]{0.32\columnwidth}
     \centering
     \includegraphics[width=\linewidth]{modo_methodology/Cancelation_Situation.pdf}
    %  \vspace*{-3mm}
     {\\(b) Anticipated vehicle return}
    \end{minipage}
    %\hspace{1cm}
    %\hspace{5mm}
    \begin{minipage}[b]{0.32\columnwidth}
    % \vspace{5mm}
     \centering
     \includegraphics[width=\linewidth]{modo_methodology/Consecutive_without_legend.pdf}
     {\\(c) Extended travel situation}
    \end{minipage}
    \caption{Possible vehicle status during the Modo crawling. In (a) a normal booking and usage situation; (b) a cancellation situation; (c) a consecutive booking situation.}
    \label{fig:capturas}
\end{figure}

Besides, we also collect base stations location, vehicle models and whether the vehicle is electric or hybrid. 
Table~\ref{table:dataModo} summarizes the data we have collected from Modo. We stored 134 millions of records in 5 months, from a fleet of 682 vehicles distributed in 528 stations, each of them with one or more cars. The stations are located in Vancouver, Canada, and its neighbor cities. This data allows us to analyze more than %149 thousands of booking records and 
98\,000 travels.\footnote{Data are available on http://netlab.ice.ufjf.br/index.php/carsharingdata/}
% \footnote{Data will be available for researchers upon request.}

\begin{table}[tbh]
	\setlength{\tabcolsep}{2.3pt}
	\centering
		\begin{tabular}{llr}
		%\multicolumn{2}{l}{Description}                     & Amount \\ 
		\hline
		\multicolumn{2}{l}{\# of Collected Records}  & $\approx$ 134\,000\,000\\
		\multicolumn{2}{l}{\# of Booking Records} & 149\,732  \\
		\multicolumn{2}{l}{\# of Travels Records} & 98\,915   \\
		\multicolumn{2}{l}{\# of Stations} & 528    \\\hline   
		\multirow{3}{*}{\# of Vehicles}       & - Common    & 530 \\
		                                      & - Hybrids  & 148 \\
		                                      & - Electrical & 4 \\
		                                      \hline
		\end{tabular}
	\caption{Summary of the Modo dataset.}
	\label{table:dataModo}
\end{table}

@@

Evo does not offer a public API to researchers. For this reason, we collect data which is publicly available at its web portal. Minute by minute, we retrieve a list of all system vehicles. Moreover, we request service snapshots, describing which vehicles are parked, where they are parked and if they are available to travel. 
We process all snapshots of the system to infer the moments a vehicle is busy (rented) or idle (parked at a station). During a snapshot, if a vehicle is listed among the system vehicles but it is not parked at any station, we infer it is in use. Then, we set-up the travel starting point as the last station the vehicle was parked. Analogously, the travel ending point will be the next station the vehicle appears in a future snapshot. The total travel time is accounted for as the difference between these snapshots times. 
For each travel we identify, we also record the end-to-end path, according to the Google Maps API. In this way, we are also able to calculate the estimated travel, taking into account the local traffic conditions. Clearly, this estimation does not take into account the car-sharing client behaviour and, as a consequence, differ from the real travel time we also store. 
One may reserve a car in Evo and cancel this reservation, within a thirty minutes range, without any charges. Thus, we infer the number of cancellation in Evo by filtering short travels (i.e., $<$ 30 minutes) where the start and end points are the same. To accommodate GPS imprecision, we consider a 3~meters threshold. 
Table~\ref{table:dataEvo} summarizes the data we collect from Evo.
%In short, we collected more than 10 million records and 1 million travels, of a fleet with more than a one thousand vehicles and 130 stations. 
Note that this service does not need a large number of stations because the user can park the car in some public park spots in the service area, that is called home zone (Vancouver and its neighbor cities).

@@

\begin{tabular}{llr}
%\multicolumn{2}{l}{Description} & Amount\\ 
\hline
\multicolumn{2}{l}{\# of Collected Records} & 142\,853\,500
\\
\multicolumn{2}{l}{\# of Travels Records} &  644\,887\\
%1\,232\,262\\
\multicolumn{2}{l}{\# of Stations}  & 130  \\\hline
\multicolumn{2}{l}{\# of Vehicles} & 1\,237
%1\,003 
\\\hline
\end{tabular}
\caption{Summary of the Evo data collection.}
\label{table:dataEvo}
\end{table}

@@

Car2Go offers APIs providing information about available cars at the moment of the request. Each API request returns, among other information, the car unique ID, its position and other fields which specifically describe the car status. Therefore the API response is semantically equivalent to the Evo's one. In this way, we applied the same methodology to gather and store the Car2go data too.

There are two main events, which changes the car status, clearly observable from the data. Considering the current time instant $t_i$: 
\begin{itemize}
     \item if in $t_i$ the car is present in the API response and at time $t_{i+1}$ it is not, that car passes from available to rented. %It represents a parking event finish and a booking.
     \item if in $t_i$ the car is \emph{not} present and at time $t_{i+1}$ it reappears in the API reply, that car passes from rented  to available. It represents a booking finish and a parking beginning.  Indeed, for privacy constraints, the position of the car during a booking is not available.
\end{itemize}

Notice that from a single rented status is impossible to estimate the travelled distance: by computing the Euclidean or Haversine 
distance we obtain only a lower bound of the real travel distance which is practically too optimistic to be used as a primary travel estimation. To improve this estimation we attach to each entry the distance provided by the Google Maps API. 
As in Evo's methodology, we infer the number of cancellations by filtering short travels where the start and end points are very close. Table~\ref{table:dataCar2Go} summarizes Car2go dataset. We have more than one million travels in our thirteen months of data. As a free-floating service, Car2Go does not have stations but it has an operation zone, that covers a large area of Vancouver city and North-Vancouver.

\begin{table}[htb]
	\centering
	\setlength{\tabcolsep}{2.3pt}
	\begin{tabular}{llr}
	%\multicolumn{2}{l}{Description} & Amount\\ 
	\hline
	\multicolumn{2}{l}{\# of Travels Records} & 1\,095\,577\\
	%1\,340\,053\\
	\multicolumn{2}{l}{\# of Vehicles} & 1\,077
	%1\.273 
	\\\hline
	\end{tabular}
	\caption{Summary of the Car2Go data collection.}
	\label{table:dataCar2Go}
\end{table}

>>>>

%MODO data collection
The Modo data collection process was conducted with a crawler that uses its public API. 
First, we request to the Modo API the list of all vehicles of the service. 
Then, minute by minute, we request the status of each of these vehicles. Each request returns the schedule of a vehicle, informing the periods it will be available for the next 24-hours. Moreover, it returns the vehicle location, i.e., the station with its identifier. 
Note that Modo API does not return specific vehicle status, nor any information that could be used to identify users of the system. We uncover if a vehicle is busy or idle based on its reservation period and the current observation time. In other words, we collect several vehicle schedules and compare each other. Figure~\ref{fig:4_3_capturas} illustrates the process of collecting data for a given vehicle. Each data sample corresponds to a request to the API in the order they occur. Data sample \#1 is the result of the API request at minute 1 ($t=1$), data sample \#2 is the result of the API request at minute 2 ($t=2$), etc. At each data sample, the blue dot represents the time a vehicle will be available. We highlight three possible situations:

\begin{itemize}
\item First, as shown in figure~\ref{fig:4_3_capturas}(a), at $t=1$ a given vehicle is shown reserved up to $t=5$.  At $t=2$, the new request to the Modo API still show us that the vehicle will be available only at $t=5$. Each of the following requests to the API confirms the booking period. At the time $t=6$, we perform a request to the API and the vehicle is no longer booked. In sum, we are able to infer that someone booked the vehicle before or at $t=1$, and returned it to the station at $t=5$.

\item Second, as shown in figure~\ref{fig:4_3_capturas}(b), at $t=1$  the Modo API returns that a given vehicle is reserved up to $t=6$.  However, in this case, a request at $t=5$ shows the vehicle is no longer reserved. In this case, we can infer that the user returned the vehicle earlier to the station which means she/he used the vehicle only up to $t=5$.

\item Finally, as shown in figure~\ref{fig:4_3_capturas}(c), the user may extend the booking period. More precisely, at $t=1$ the given vehicle is reserved up to $t=5$. At the third request, we note that the vehicle will no longer be available at $t=5$ but $t=6$. The following API requests confirm the use of the car until $t=6$.
\end{itemize}

@@

    \begin{minipage}[b]{0.32\columnwidth}
    \centering
     \includegraphics[width=\linewidth]{modo_methodology/Normal_Situation.pdf}
     {\\(a) Normal \\situation}
    \end{minipage}
    % \hspace{5mm}
    \begin{minipage}[b]{0.32\columnwidth}
     \centering
     \includegraphics[width=\linewidth]{modo_methodology/Cancelation_Situation.pdf}
    %  \vspace*{-3mm}
     {\\(b) Anticipated vehicle return}
    \end{minipage}
    %\hspace{1cm}
    %\hspace{5mm}
    \begin{minipage}[b]{0.32\columnwidth}
    % \vspace{5mm}
     \centering
     \includegraphics[width=\linewidth]{modo_methodology/Consecutive_without_legend.pdf}
     {\\(c) Extended travel situation}
    \end{minipage}
    \caption{Possible vehicle status during the Modo crawling. In (a) a normal booking and usage situation; (b) a cancellation situation; (c) a consecutive booking situation.}
    \label{fig:4_3_capturas}
\end{figure}

Besides, we also collect base stations location, vehicle models and whether the vehicle is electric or hybrid. 
Table~\ref{table:4_3_dataModo} summarizes the data we have collected from Modo. We stored 134 millions of records in 5 months, from a fleet of 682 vehicles distributed in 528 stations, each of them with one or more cars. The stations are located in Vancouver, Canada, and its neighbor cities. This data allows us to analyze more than %149 thousands of booking records and 
98\,000 travels.\footnote{Data are available on http://netlab.ice.ufjf.br/index.php/carsharingdata/}
% \footnote{Data will be available for researchers upon request.}

\begin{table}[tbh]
	\setlength{\tabcolsep}{2.3pt}
	\centering
		\begin{tabular}{llr}
		%\multicolumn{2}{l}{Description}                     & Amount \\ 
		\hline
		\multicolumn{2}{l}{\# of Collected Records}  & $\approx$ 134\,000\,000\\
		\multicolumn{2}{l}{\# of Booking Records} & 149\,732  \\
		\multicolumn{2}{l}{\# of Travels Records} & 98\,915   \\
		\multicolumn{2}{l}{\# of Stations} & 528    \\\hline   
		\multirow{3}{*}{\# of Vehicles}       & - Common    & 530 \\
		                                      & - Hybrids  & 148 \\
		                                      & - Electrical & 4 \\
		                                      \hline
		\end{tabular}
	\caption{Summary of the Modo dataset.}
	\label{table:4_3_dataModo}
\end{table}

@@

Evo does not offer a public API to researchers. For this reason, we collect data which is publicly available at its web portal. Minute by minute, we retrieve a list of all system vehicles. Moreover, we request service snapshots, describing which vehicles are parked, where they are parked and if they are available to travel. 
We process all snapshots of the system to infer the moments a vehicle is busy (rented) or idle (parked at a station). During a snapshot, if a vehicle is listed among the system vehicles but it is not parked at any station, we infer it is in use. Then, we set-up the travel starting point as the last station the vehicle was parked. Analogously, the travel ending point will be the next station the vehicle appears in a future snapshot. The total travel time is accounted for as the difference between these snapshots times. 
For each travel we identify, we also record the end-to-end path, according to the Google Maps API. In this way, we are also able to calculate the estimated travel, taking into account the local traffic conditions. Clearly, this estimation does not take into account the car-sharing client behaviour and, as a consequence, differ from the real travel time we also store. 
One may reserve a car in Evo and cancel this reservation, within a thirty minutes range, without any charges. Thus, we infer the number of cancellation in Evo by filtering short travels (i.e., $<$ 30 minutes) where the start and end points are the same. To accommodate GPS imprecision, we consider a 3~meters threshold. 
Table~\ref{table:4_3_dataEvo} summarizes the data we collect from Evo.
%In short, we collected more than 10 million records and 1 million travels, of a fleet with more than a one thousand vehicles and 130 stations. 
Note that this service does not need a large number of stations because the user can park the car in some public park spots in the service area, that is called home zone (Vancouver and its neighbor cities).

@@

\begin{tabular}{llr}
%\multicolumn{2}{l}{Description} & Amount\\ 
\hline
\multicolumn{2}{l}{\# of Collected Records} & 142\,853\,500
\\
\multicolumn{2}{l}{\# of Travels Records} &  644\,887\\
%1\,232\,262\\
\multicolumn{2}{l}{\# of Stations}  & 130  \\\hline
\multicolumn{2}{l}{\# of Vehicles} & 1\,237
%1\,003 
\\\hline
\end{tabular}
\caption{Summary of the Evo data collection.}
\label{table:4_3_dataEvo}
\end{table}

@@

The car2go data collection is widely described in chapter \ref{chap:2_datacollection}. 
%In a nutshell, car2Go offers APIs providing information about available cars at the moment of the request. Each API request returns, among other information, the car unique ID, its position and other fields which specifically describe the car status. Therefore the API response is semantically equivalent to the Evo's one. In this way, we applied the same methodology to gather and store the Car2go data too.

Recalling that the software is able to detect two events, corresponding to the car status change, clearly observable from the data. Indeed considering the current time instant $t_i$: 
\begin{itemize}
     \item if in $t_i$ the car is present in the API response and at time $t_{i+1}$ it is not, that car passes from available to rented. %It represents a parking event finish and a booking.
     \item if in $t_i$ the car is \emph{not} present and at time $t_{i+1}$ it reappears in the API reply, that car passes from rented  to available. It represents a booking finish and a parking beginning.  Indeed, for privacy constraints, the position of the car during a booking is not available.
\end{itemize}

%Notice that from a single rented status is impossible to estimate the travelled distance: by computing the Euclidean or Haversine 
%distance it is possible to deduce a lower bound of the real travel distance which is practically too optimistic to be used as a primary travel estimation. To improve this estimation we attach to each entry the distance provided by the Google Maps API. 
%As in Evo's methodology, we infer the number of cancellations by filtering short travels where the start and end points are very close.
Notice that, the straight distance (computed with the Haversine formula) between the ride starting and final points it is not the real driven distance hidden for privacy issues. For this reason the software attaches for each ride the distance provided by the Google Maps API, in order have a better estimation of the driven pattern.

 Table~\ref{table:4_3_dataCar2Go} summarizes Car2go dataset. We have more than one million travels in thirteen months of data. As a free-floating service, Car2Go does not have stations but it has an operation zone, that covers a large area of Vancouver city and North-Vancouver.

\begin{table}[htb]
	\centering
	\setlength{\tabcolsep}{2.3pt}
	\begin{tabular}{llr}
	%\multicolumn{2}{l}{Description} & Amount\\ 
	\hline
	\multicolumn{2}{l}{\# of Travels Records} & 1\,095\,577\\
	%1\,340\,053\\
	\multicolumn{2}{l}{\# of Vehicles} & 1\,077
	%1\.273 
	\\\hline
	\end{tabular}
	\caption{Summary of the Car2Go data collection.}
	\label{table:4_3_dataCar2Go}
\end{table}
